---
title: "Comparing the Daily Schedules in the NHTS from 2009 and 2017"
author: "Greg Macfarlane"
date: "4/9/2018"
output: html_document
---

```{r setup, include=FALSE, cache = FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
library(tidyverse)
library(lubridate)
```

# Abstract
Travel modelers commonly use passive transportation data, such as cellular or
GPS origin-destination matrices, to calibrate or validate their planning models,
expanding and adjusting behavioral models estimated from a small-sample local
household survey.  Recently, there has been interest in deriving synthetic
records of individual travel directly from the passive data, with the support of
other datasets. We previously developed a method (NCHRP IDEA Project 184) to
build synthetic daily activity and travel patterns for a complete population. In
this method, simulated individuals used the 2009 NHTS as a basis for tour
patterns while passive and location-specific origin-destination matrices
spatially locate the simulated tours within a specific region. 

In this work, we update our tour database to the 2017 NHTS and comment on
observed differences in the resulting simulated travel patterns. Specifically,
we will consider the temporal distribution of weekday trips: the NHTS in 2009
and 2017 both show a higher proportion of mid-day, off-peak travel than other
comparable local household surveys. We will examine the consequences of this for
temporal activity patterns and in traffic assignment, and consider heterogeneity
in distribution by city size, geographic region, and traveler attributes such as
employment status and vehicle availability.


# Background

In a conventional travel modeling paradigm, planners estimate econometric models
of travel behavior --- mode choice, destination choice, etc. --- and expand the
survey to match the total population distribution. Under this paradigm, a local
dataset on which to estimate the models is important to ensure the models
capture region-specific behavior. Passively collected travel data such as
cellular phone traces and highway counts are used to validate and calibrate the
models, but are not useful in estimation because they do not contain information
on the behavior and choices of individuals.

The general National Household Travel Survey (NHTS) is not typically suitable
for a conventional travel modeling exercise, because the local sampling rate is
too small and the geographic resolution is too large for model estimation or
calibration (with some modifications, state or MPO-specific NHTS add-ons may be
used).

In the past several years, there have been efforts in the travel demand modeling
community to explore the use of so-called "data-driven" travel demand models. 
In contrast to conventional models, a data-driven model uses the large-scale
passive data explicitly in the model process. We previously described a
methodology for such a model in NCHRP IDEA Project 184; in this methodology,
synthetic individuals generate tours by referencing the NHTS for individuals
living in similarly-sized cities and then draw locations for their tour
activities from origin-destination datasets specific to the model region. The
resulting synthetic tours can then be assigned to highway and transit networks
to understand route and mode choice. In this research we use the open-source
transport network simulation MATSim. 

The most recent two iterations of the NHTS were in 2009 and 2017. Though the
methodologies between successive survey collections can change somewhat, enough
similarities exist that a comparison of travel trends between is useful and
important. In this research we consider the diurnal distribution of tour start
times and trips-in-motion reported in the NHTS and other related household
surveys, and compare the results to the data-driven model developed by Transport
Foundry.

# Methodology and Results

Transport Foundry has a data-driven model for Asheville, North Carolina, a 
medium-sized city in western North Carolina. The CBSA population is approximately
400,000 people. In this section we compare the distribution of tours and
trips-in-motion in the simulation against the public NHTS schedules data from
which model is derived.


### Tour start distribution

Beginning with the public data files for the 2009 and 2017 NHTS for respondents
living in an MSA of between 250,000 and 500,000 people, we calculated
the probability of archetype tours beginning each in each 90 minute period. An
archetype tour is means of tour classification: a "simple work" tour archetype
includes a single work activity, a "multi-part work" tour includes more than one
activity, and a "composite" tour includes both work and non-work activities.
The distribution of these probabilities by time of day are given in the plot
below.

```{r schedules, fig.cap="Archetype tour probability for workers by time of day. The balance of the probability is the probability that no tour will start in the period."}
schedules <- bind_rows(
  read_csv("data/02_2009_schedule_next_tour.csv") %>% mutate(year = 2009),
  read_csv("data/02_2017_schedule_next_tour.csv") %>% mutate(year = 2017)
) %>%
  mutate(
    time = citycastr:::get_military_time(bin_start),
    time = lubridate::as_datetime(
      stringr::str_c("2018-01-01 ", substr(time,1,2), ":", substr(time, 3, 4), ":00"))
  )

ggplot(schedules %>% filter(tour_type != "No tour", worker), 
       aes(y = prob, x = time, fill = tour_type)) +
  geom_col() + 
  facet_wrap(~year) +
  xlab("Clock Time (24 hr)") + ylab("Tour Start Probability") +
  scale_y_continuous(labels = scales::percent) +
  scale_x_datetime(labels = scales::date_format("%H:%M"))
```

From this plot, it appears that workers in the 2017 survey are somewhat less
likely to make tours that involve a work activity.

That is, the largest increase in
probabilities are for multi-part and simple work tours, and the decrease
has come primarily from simple work and composite from-work tours. This may 
indicate a true trend in travel behavior, or a discrepany in how respondents
classified the purpose of their activities, or simply a random variation. 
That said, there are over 20,000 respondents to the NHTS in cities of this
size in both survey years, so a random variation is unlikely.

With this information as a background, we now turn to understanding how this
change in tour information affects the simulated trips and tours in the 
data-driven model. To do this, we consider the temporal distribution of 
trips-in-motion.

### Trips-in-Motion
Using the NHTS data for cities between 250,000 and 500,000 people, we calculated
the weighted trips-in-motion for 15 minute intervals throughout a 24-hour
weekday period. To be included in a period, a trip either needs to be currenly
underway or completed entirely within the next 15 minutes. We generated separate
distributions for home-based work (HBW) trips and all trips.

```{r read_nhts}
source("R/clean_nhts_trips.R")
trips09 <- read_csv("nhts2009/nhts_day.csv.gz") %>% clean_nhts_trips("2009")
trips17 <- read_csv("nhts2017/trippub.csv.gz") %>% clean_nhts_trips("2017")
```

```{r tim_nhts}
trips <- bind_rows(trips17, trips09) %>%
  filter(!weekend) %>%
  filter(msasize == "250k - 499k")

tim_nhts <- build_trips_in_motion(trips, bin_size = 15)
tim_all_purposes <- tim_nhts %>%
  group_by(dataset, time) %>%
  summarise(
    trips = sum(trips),
    purpose = "All"
  )
tim_nhts <- bind_rows(tim_nhts %>% filter(purpose == "HBW"), tim_all_purposes) %>%
  ungroup() 
```


We also obtained comparable distributions from two other data sources: a local 
household travel survey for the Hickory, North Carolina region and a 2009 NHTS
add-on for the Blacksburg, Virginia region. Though these regions are both smaller
than Asheville in terms of population, they share common elements in culture,
geography, and infrastructure. The plot below shows the distribution of these
trips-in-motion in each dataset. All the data show a pattern typical of daily
travel: there are peaks in the AM and PM representing the beginning and end of
the work day. The peaks are somewhat more pronounced in the Hickory survey but
overall the four datasets are remarkably similar.


```{r small_tim}
tim_small <- bind_rows(
  read_csv("data/hickory_tod.csv", col_types = list(time = col_character())) %>%
    mutate(dataset = "Hickory"),
  read_csv("data/blacksburg_tod.csv", col_types = list(time = col_character())) %>%
    mutate(dataset = "Blacksburg (NHTS Add-On)")
) %>%
  separate(time, c("hours", "minutes"), fill = "right") %>%
  mutate(
    minutes = ifelse(is.na(minutes), 0, minutes),
    minutes = ifelse(minutes == 3, 30, minutes),
    time = lubridate::ymd_hm(paste0("2018-01-01 ", hours, ":", minutes))
  ) %>%
  select(time, All = AllTrips, HBW = HBWTrips, dataset) %>%
  gather(purpose, trips, All, HBW)

  
```

```{r trips_in_motion_plot, fig.cap=""}
trips_in_motion <-  bind_rows(tim_small, tim_nhts) %>%
  group_by(dataset, purpose) %>%
  mutate(
    trips = trips / sum(trips)
  )

ggplot(trips_in_motion, aes(x = time, y = trips, color = dataset)) +
  geom_line() +
  scale_color_discrete(name = "Dataset") +
  facet_wrap(~purpose) +
  xlab("Local Time") + ylab("Percent of Daily Trips") +
  scale_y_continuous(labels = scales::percent) +
  scale_x_datetime(labels = scales::date_format("%H:%M"))
```


The data-driven model generates synthetic travel demand in two stages. First, 
the model generates an initial demand considering the tour patterns and
origin-destination data. This initial demand is then fed into MATSim, which 
adjusts the activity start and end times, the travel routes, and the travel mode
in an attempt to iteratively improve the utility of each person's day. Travel 
time and cost is minimized, and time spent doing activities is maximized.


```{r ash_demand, eval=FALSE}
ash17_demand <- read_csv("data/asheville_2017_demand.csv") 
ash09_demand <- read_csv("data/asheville_2009_demand.csv") 
ash17_assign <- read_csv("data/asheville_2017_supply.csv")
ash09_assign <- read_csv("data/asheville_2009_supply.csv")
```

We ran the data-driven model for a synthetic population in Asheville, North
Carolina using the 2009 and 2017 NHTS as a schedules database. Because the 
simulation for a city includes far more records than the NHTS responses, we 
calculate the trips-in-motion for 60 minute periods instead of 15-minute
periods.

```{r nhts_60}
tim_nhts_60 <- build_trips_in_motion(trips, bin_size = 60)
tim_all_purposes_60 <- tim_nhts_60 %>%
  group_by(dataset, time) %>%
  summarise(
    trips = sum(trips),
    purpose = "All"
  )
tim_nhts_60 <- bind_rows(tim_nhts_60 %>% filter(purpose == "HBW"), tim_all_purposes_60) %>%
  ungroup() 
```


```{r tim_ash_demand, eval = FALSE}
tim_demand <- bind_rows(
  build_tim_from_demand(ash17_demand, bin_size = 60) %>% mutate(dataset = "2017 Asheville"),
  build_tim_from_demand(ash09_demand, bin_size = 60) %>% mutate(dataset = "2009 Asheville")
)

tim_all_demand <- tim_demand %>%
  group_by(dataset, time) %>%
  summarise(
    trips = sum(trips),
    purpose = "All"
  )

tim_demand <- bind_rows(tim_demand %>% filter(purpose == "HBW"), tim_all_demand) %>%
  ungroup() 
  
```

```{r ash_supply, eval=FALSE}
tim_supply <- bind_rows(
  build_tim_from_supply(ash17_assign, bin_size = 15) %>% mutate(dataset = "2017 Asheville - Replanned"),
  build_tim_from_supply(ash09_assign, bin_size = 15) %>% mutate(dataset = "2009 Asheville - Replanned")
)

tim_all_supply <- tim_supply %>%
  group_by(dataset, time) %>%
  summarise(
    trips = sum(trips),
    purpose = "All"
  )

tim_supply <- bind_rows(tim_supply %>% filter(purpose == "HBW"), tim_all_supply) %>%
  ungroup() 
```

```{r tim_supply_plot, eval = FALSE}
trips_in_motion <-  bind_rows(tim_nhts, tim_supply) %>%
  group_by(dataset, purpose) %>%
  mutate(
    trips = trips / sum(trips)
  ) %>%
  separate(
    dataset, c("year", "dataset"), extra = "merge"
  )

ggplot(trips_in_motion, aes(x = time, y = trips, color = dataset)) +
  geom_line() +
  scale_color_discrete(name = "Dataset") +
  facet_wrap(year~purpose) +
  xlab("Local Time") + ylab("Percent of Daily Trips") +
  scale_y_continuous(labels = scales::percent) +
  scale_x_datetime(labels = scales::date_format("%H:%M"))
```


### Acknowledgements
We are grateful to Kyle Ward of WSP who provided trips-in-motion data for
Hickory and Blacksburg.