---
title: "Comparing the Daily Schedules in the NHTS from 2009 and 2017"
author: "Greg Macfarlane PhD PE and Josie Kressner PhD, Transport Foundry"
date: "8 August 2018"
output:
  bookdown::pdf_document2:
    toc: false
references:
- id: trbidea
  title: Synthetic Household Travel Data Using Consumer and Mobile Phone Data
  author:
  - family: Kressner
    given: Josephine
  container-title: NCHRP IDEA Program
  URL: http://www.trb.org/Main/Blurbs/176216.aspx
  issue: 184
  publisher: Transportation Research Board
  issued:
    year: 2017
    month: 3
- id: nhts09
  author:
  - family: U.S. Department of Transportation
  title: National Household Travel Survey
  URL: https://nhts.ornl.gov
  issued:
    year: 2017
---

```{r setup, include=FALSE, cache = FALSE}
knitr::opts_chunk$set(echo = FALSE, cache = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(lubridate)
```

# Abstract
Travel modelers commonly use passive transportation-related data, such as
cellular or GPS origin-destination matrices, to calibrate or validate their
planning models. The passive data are generally used, in this case, to expand
and adjust behavioral models estimated from a small-sample local household
survey. Recently, there has been interest in deriving synthetic records of
individual travel directly from the passive data, with the support of other
datasets. @trbidea previously developed a method to build
synthetic daily activity and travel patterns for a complete population. In this
method, simulated individuals used the 2009 NHTS as a basis for tour patterns
while passive origin-destination matrices spatially locate the simulated tours
within a specific region.

In this work, we update our tour patterns to the 2017 NHTS and comment on
observed differences in the resulting simulated travel patterns. Specifically,
we consider the temporal distribution of weekday trips. We also examine the
consequences of these differences in the synthesized daily schedules and in
their traffic assignment. The results show that the NHTS has a similarly high
proportion of mid-day, off-peak travel in both 2009 and 2017, and that this
proportion is higher than other comparable local household surveys. The results
also show that the change from the 2009 to the 2017 dataset does not
substantively affect the simulated demand or assignment in a data-driven travel
model using NHTS data.


# Background

In a conventional travel modeling paradigm, planners estimate econometric models
of travel behavior --- mode choice, destination choice, etc. --- and expand the
survey to match the total population distribution. Under this paradigm, a local
dataset on which to estimate the models is important to ensure the models
capture region-specific behavior. Passively collected travel data such as
cellular phone traces and highway counts are used to validate and calibrate the
models, but are not useful in estimation because they do not contain information
on the behavior and choices of individuals.

The general National Household Travel Survey (NHTS)[@nhts09] is not typically suitable
for a conventional travel modeling exercise, because the local sampling rate is
too small and the geographic resolution is too large for model estimation or
calibration (with some modifications, state or MPO-specific NHTS add-ons may be
used).

In the past several years, there have been efforts in the travel demand modeling
community to explore the use of so-called "data-driven" travel demand models.
In contrast to conventional models, a data-driven model uses the large-scale
passive data explicitly in the model process. @trbidea previously described a
methodology for such a model; in this methodology,
synthetic individuals generate tours by referencing the NHTS for individuals
living in similarly-sized cities and then draw locations for their tour
activities from origin-destination datasets specific to the model region. The
resulting synthetic tours can then be assigned to highway and transit networks
to understand route and mode choice. In this research we use an open-source
transport network simulator, MATSim.

The most recent two iterations of the NHTS were in 2009 and 2017. Though the
methodologies between successive survey collections did change somewhat, enough
similarities exist that a comparison of travel trends between is useful and
important. In this research we consider the diurnal distribution of tour start
times and trips-in-motion reported in the NHTS and other related household
surveys, and we compare the results to the data-driven model developed by
Transport Foundry.


# Methodology
We developed a data-driven model for Asheville, North Carolina, which is a
medium-sized city in western North Carolina. The core-based statistical area
(CBSA) in Asheville has a  population of approximately 400,000 people. We used
the public data files for the 2009 and 2017 NHTS for respondents living in a
CBSA with a population between 250,000 and 500,000 people. In both years, there
are more than 20,000 respondents in cities of this size.

We compare the distribution of tour types and trips-in-motion in the data-driven
simulation against the public NHTS schedules data from which the model is
derived.


# Results
## Tour Start Distribution
For the distribution of tour types, we calculate the probability of archetype
tours beginning in each 90 minute period for workers only, accounting for person
weights in the NHTS data. Non-workers are excluded in this plot for simplicity.
An archetype tour is a means of tour classification: a "simple work" tour
archetype includes a single work activity, a "multi-part work" tour includes
more than one work activity, and a "composite" tour includes both work and
non-work activities.

The distribution of probabilities for each archetype of tours by time of day are
given in Figure \@ref(fig:schedules). A distribution of the difference between the
distributions by purpose and period is in \@ref(fig:spread-schedules). From this
plot, it appears that workers in the 2017 survey are somewhat less likely to
make tours that involve a work activity, and somewhat more respondents take
multi-part work tours. There is also less of a peak in the late afternoon in
2017. Overall however, the distributions are comparable.


```{r schedules, fig.cap="Probability of an archetype tour starting for workers by time of day.", dev = 'tikz'}
schedules <- bind_rows(
  read_csv("data/02_2009_schedule_next_tour.csv") %>% mutate(year = 2009),
  read_csv("data/02_2017_schedule_next_tour.csv") %>% mutate(year = 2017)
) %>%
  mutate(
    time = citycastr:::get_military_time(bin_start),
    time = lubridate::as_datetime(
      stringr::str_c("2018-01-01 ", substr(time,1,2), ":",
                     substr(time, 3, 4), ":00")),
    tour_type = gsub("Composite", "Mixed", tour_type),
    tour_type = gsub("to and from", "to \\\\& from", tour_type)
  ) %>%
  filter(tour_type != "No tour")

ggplot(schedules %>% filter(worker),
       aes(y = prob, x = time)) +
  geom_col(aes(fill = tour_type)) +
  facet_wrap(~year)  +
  scale_fill_discrete(name = "") +
  xlab("Clock Time (24 hr)") + ylab("Tour Start Probability") +
  scale_y_continuous(labels = scales::percent_format(suffix = "\\%")) +
  scale_x_datetime(labels = scales::date_format("%H:%M")) +
  tfplotr::theme_tf()
```


```{r spread-schedules, fig.cap="Difference in archetype tour start probability between 2017 and 2009 NHTS for workers by time of day.", dev = 'tikz'}
spread_schedules <-  schedules %>%
  spread(year, prob, fill = 0) %>%
  transmute(
    worker, tour_type, time,
    diff = `2017` - `2009`)

all_schedules <- spread_schedules %>%
  group_by(worker, time) %>%
  summarise(diff = sum(diff)) %>%
  mutate(tour_type = "All tours")

ggplot(bind_rows(spread_schedules, all_schedules) %>% filter(worker),
       aes(y = diff, x = time)) +
  geom_col() +
  facet_wrap(~tour_type) +
  xlab("Clock Time (24 hr)") + ylab("Tour Start Probability") +
  scale_y_continuous(labels = scales::percent_format(suffix = "\\%")) +
  scale_x_datetime(labels = scales::date_format("%H:%M")) +
  tfplotr::theme_tf()
```


That is, the largest increase in
probabilities are for multi-part and simple work tours, and the decrease
has come primarily from simple work and composite from-work tours. This may
indicate a true trend in travel behavior, or a discrepany in how respondents
classified the purpose of their activities, or simply a random variation.
That said, there are over 20,000 respondents to the NHTS in cities of this
size in both survey years, so a random variation is unlikely.

With this information as a background, we now turn to understanding how this
change in tour information affects the simulated trips and tours in the
data-driven model. To do this, we consider the temporal distribution of
trips-in-motion.

## Trips-in-Motion
For a distribution of trips-in-motion, we calculate the number of trips moving
during each 15 minute interval throughout a 24-hour weekday period. To be
counted in a 15 minute interval, a trip either needs to be currently underway or
completed entirely within the next 15 minutes. We generate separate
distributions for home-based work (HBW) trips and all trips.

```{r read_nhts}
source("R/clean_nhts_trips.R")
trips09 <- read_csv("nhts2009/nhts_day.csv.gz") %>% clean_nhts_trips("2009")
trips17 <- read_csv("nhts2017/trippub.csv.gz") %>% clean_nhts_trips("2017")
```

```{r tim_nhts}
trips <- bind_rows(trips17, trips09) %>%
  filter(!weekend) %>%
  filter(msasize == "250k - 499k")

tim_nhts <- build_trips_in_motion(trips, bin_size = 15)
tim_all_purposes <- tim_nhts %>%
  group_by(dataset, time) %>%
  summarise(
    trips = sum(trips),
    purpose = "All"
  )
tim_nhts <- bind_rows(tim_nhts %>% filter(purpose == "HBW"), tim_all_purposes) %>%
  ungroup()
```

We also obtained comparable distributions from two other data sources: a local
household travel survey for the Hickory, North Carolina region and a 2009 NHTS
add-on for the Blacksburg, Virginia region. Though these regions are both
smaller than Asheville in terms of population, they share common elements in
culture, geography, and infrastructure. Figure \@ref(fig:tim) shows the
distribution of these trips-in-motion in each dataset. All the data show a
pattern typical of daily travel: there are peaks in the AM and PM representing
the beginning and end of the work day. The peaks are somewhat more pronounced in
the Hickory survey, but overall the four datasets are remarkably similar.


```{r small_tim}
tim_small <- bind_rows(
  read_csv("data/hickory_tod.csv", col_types = list(time = col_character())) %>%
    mutate(dataset = "Hickory"),
  read_csv("data/blacksburg_tod.csv", col_types = list(time = col_character())) %>%
    mutate(dataset = "Blacksburg (NHTS Add-On)")
) %>%
  separate(time, c("hours", "minutes"), fill = "right") %>%
  mutate(
    minutes = ifelse(is.na(minutes), 0, minutes),
    minutes = ifelse(minutes == 3, 30, minutes),
    time = lubridate::ymd_hm(paste0("2018-01-01 ", hours, ":", minutes))
  ) %>%
  select(time, All = AllTrips, HBW = HBWTrips, dataset) %>%
  gather(purpose, trips, All, HBW)


```

```{r tim, fig.cap="Distribution of trips-in-motion for NHTS for medium-sized cities and comparison datasets.", dev = 'tikz'}
trips_in_motion <-  bind_rows(tim_small, tim_nhts) %>%
  group_by(dataset, purpose) %>%
  mutate(
    trips = trips / sum(trips)
  )

ggplot(trips_in_motion, aes(x = time, y = trips, color = dataset)) +
  geom_line() +
  scale_color_discrete(name = "Dataset") +
  facet_wrap(~purpose) +
  xlab("Local Time") + ylab("Percent of Daily Trips") +
  scale_y_continuous(labels = scales::percent_format(suffix = "\\%")) +
  scale_x_datetime(labels = scales::date_format("%H:%M")) +
  tfplotr::theme_tf()
```

The data-driven model generates synthetic travel demand in two stages. First,
the model generates an initial demand considering the tour patterns and
origin-destination data. This initial demand is then fed into MATSim, which
adjusts the activity start and end times, the travel routes, and the travel mode
in an attempt to account for congestion (considering all other people's travel)
and to iteratively improve the utility of each person's day. Travel time and
cost is minimized, and time spent doing activities is maximized.

```{r ash_demand}
ash17_demand <- read_csv("data/asheville_2017_demand.csv")
ash09_demand <- read_csv("data/asheville_2009_demand.csv")
ash17_assign <- read_csv("data/asheville_2017_supply.csv")
ash09_assign <- read_csv("data/asheville_2009_supply.csv")
```

We ran the data-driven model to generate travel demand for a synthetic
population in Asheville, North Carolina using the 2009 and 2017 NHTS as a
tour archetype database and subsequently ran the routing and replanning
simulation in MATSim. Figure \@ref(fig:tim-supply) shows the distribution of
trips-in-motion in 15 minute bins.

```{r tim_ash_demand}
tim_demand <- bind_rows(
  build_tim_from_demand(ash17_demand, bin_size = 15) %>% mutate(dataset = "2017 Asheville (Demand)"),
  build_tim_from_demand(ash09_demand, bin_size = 15) %>% mutate(dataset = "2009 Asheville (Demand)")
)

tim_all_demand <- tim_demand %>%
  group_by(dataset, time) %>%
  summarise(
    trips = sum(trips),
    purpose = "All"
  )

tim_demand <- bind_rows(tim_demand %>% filter(purpose == "HBW"), tim_all_demand) %>%
  ungroup()

```

```{r ash_supply}
tim_supply <- bind_rows(
  build_tim_from_supply(ash17_assign, bin_size = 15) %>% mutate(dataset = "2017 Asheville (Replanned)"),
  build_tim_from_supply(ash09_assign, bin_size = 15) %>% mutate(dataset = "2009 Asheville (Replanned)")
)

tim_all_supply <- tim_supply %>%
  group_by(dataset, time) %>%
  summarise(
    trips = sum(trips),
    purpose = "All"
  )

tim_supply <- bind_rows(tim_supply %>% filter(purpose == "HBW"), tim_all_supply) %>%
  ungroup()
```

```{r tim-supply, fig.cap="Trips-in-motion for NHTS and simulated Asheville by time of day.", dev = "tikz"}
trips_in_motion_supply <-  bind_rows(tim_nhts, tim_demand, tim_supply) %>%
  group_by(dataset, purpose) %>%
  mutate(
    trips = trips / sum(trips)
  ) %>%
  separate(
    dataset, c("year", "dataset"), extra = "merge"
  )

ggplot(trips_in_motion_supply, aes(x = time, y = trips, color = dataset)) +
  geom_line() +
  scale_color_discrete(name = "Dataset") +
  facet_grid(year~purpose) +
  xlab("Local Time") + ylab("Percent of Daily Trips") +
  scale_y_continuous(labels = scales::percent_format(suffix = "\\%")) +
  scale_x_datetime(labels = scales::date_format("%H:%M")) +
  tfplotr::theme_tf()
```

The demand simulation initially generates a higher share of mid-day trips than
the NHTS source data, but the replanning simulation seems to re-peak the
distribution, particularly for work trips. Both the demand and replanning
simulations seem to have a longer tail in the evening, with more trips still
on the road than the NHTS data suggests.

It is worth noting that the NHTS data are considerably rougher than either
simulated outcome. This is both a function of the limited sample size of the
NHTS (20,000 respondents versus 400,000 simulated agents) as well as the
tendency for self-reporting diary respondents to round trip departures to
nearest half-hours.


# Conclusion
The change from the 2009 to the 2017 dataset does not substantively affect the
temporal distribution of tour start times by tour type, or the distribution of
vehicle trips-in-motion for medium-sized cities. Applying a data-driven
generation process can smooth out the demand distribution and lead to more
natural distributions of travel throughout the day for traffic assignment, but
additional research is needed to match peaking behavior with more complete
fidelity.

### Acknowledgements
We are grateful to Kyle Ward of WSP who provided trips-in-motion data for
Hickory and Blacksburg.

## References
